---
title: "Homework 1 Code"
author: "Christian Pascual"
date: "9/13/2018"
output: github_document
---

# Preliminary imports and preparation
```{r lib_imports }
library(tidyverse)

set.seed(1)
```

# Problem 1

Problem 1 reviews data types in R and type coercion. We'd like to create a data frame with columns containing one of the various types in R. Thanks to vectorization, we can base most of the other columns on the uniformly distributed numbers.

```{r p1_df }
# Extract df length for change later
desired_col_length = 10

various_types_df = tibble(
  uniform_nums = runif(desired_col_length, min = 0, max = 5),
  is_gt_2 = uniform_nums > 2,
  characters = character(desired_col_length),
  factors = factor(is_gt_2, labels = c("NOT GT2", "GT2"))
)
```

Type coercion is the conversion of one type to another. In cases where you attempt to do some operation, say taking the average, on a data type that is incompatible with that operation, R will try to switch the types and salvage the situation for you.

Here, I would expect `mean()` to work on the uniform sample and the boolean values. Usually, programming languages will interpret `TRUE` as 1 and `FALSE` as 0, so taking the average here will make sense.

Conversely, it will not work for the character vector or the factor vector because it doesn't make sense to average out a string or category.

```{r p1_mean_attempts }
means_df = tibble(
  # Expected to work
  uniform_mean = mean(various_types_df$uniform_nums),
  logicals_mean = mean(various_types_df$is_gt_2),
  
  # Not expected to work
  char_mean = mean(various_types_df$characters),
  factor_mean = mean(various_types_df$factors)
)
```

### Reminders for coding in R
* meaningful variable / object names
* readable code (one command per line; adequate whitespace and indentation; etc)
* clearly-written text to explain code and results
* a lack of superfluous code (e.g. no unused variables are defined)
